---
title: "How does commuting behavior vary across different socio-demographic characteristics, income levels, and infrastructural factors in France?"
author: "Mathilde VALLAT, Yesmine HACHANA"
format: pdf
editor: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
# Load knitr package
library(knitr)

# Set global chunk options
knitr::opts_chunk$set(
  echo = FALSE,    # Hide all code
  message = FALSE, # Suppress messages
  warning = FALSE  # Suppress warnings
)
```

```{r}
library(vroom)
library(here)
library(dplyr)
here::i_am("Final_Project_Commuter_behavior.Rproj")
final_dataset_cleaned2.0 <- vroom(here("final_dataset_cleaned2.0.csv"))
mob_pro_cleaned <- vroom(here("mob_pro_cleaned.csv"))
```

```{r}
# Identify columns to exclude and the multiplier column
columns_to_exclude <- c("code_epci", "intercommunal_label", "population_density_2021", "intercommunal_population_2021", "nb_stops", "median_living_standard_2021")
multiplier_column <- "intercommunal_population_2021"

# Create a new dataset where selected columns are multiplied
new_dataset <- final_dataset_cleaned2.0
columns_to_transform <- setdiff(names(final_dataset_cleaned2.0), c(columns_to_exclude, multiplier_column))

# Apply the transformation
new_dataset[columns_to_transform] <- lapply(columns_to_transform, function(col) {
  final_dataset_cleaned2.0[[col]] / 100 * final_dataset_cleaned2.0[[multiplier_column]]
})
names(new_dataset) <- gsub("share", "value", names(new_dataset))

# Choosing appropriate types
new_dataset <- new_dataset %>%
  mutate(
    # Convert to factors
    intercommunal_label = as.factor(intercommunal_label),
    
    # Ensure ID columns remain as integers
    code_epci = as.integer(code_epci),
    
    # Continuous numeric columns (already numeric, ensuring they stay so)
    across(c(intercommunal_population_2021, population_density_2021, 
             nb_stops, value_household_car, median_living_standard_2021), as.numeric),
    
    # Ensure proportion/percentage columns are numeric
    across(matches("_value$"), as.numeric)
  )

write.csv(new_dataset, "new_dataset.csv", row.names = FALSE)
```

[Link to the github
project](https://github.com/jemappelleyesmine/final_project)

# Introduction

As our modern societies have increased the distance between home and
workplace for most, commuting has become part of our daily routine. It
influences a countries energy use, urban development, and environmental
sustainability. In France, transportation accounts for a significant
share of national greenhouse gas (GHG) emissions (in 2021, transport
emissions amount to 126.0 Mt CO2 eq, equivalent to 30% of the national
total, source 1), with private vehicles being a major contributor (53%
in 2021, i.e. 66.4 Mt CO2 eq., source 1). Shifting commuters from
high-carbon modes like cars to lower-carbon options like public
transport, cycling, or walking is essential for meeting climate goals
such as those outlined in the Paris Agreement. France has committed to
reducing its carbon footprint, and understanding commuter behavior is
vital for designing effective policies to encourage sustainable
transportation choices. Beyond climate change, transport choices affect
both public health (e.g., air quality, physical activity) and quality of
life (e.g., time spent commuting, stress levels). Such that aligning
commuter preferences with sustainable goals can create co-benefits, such
as improved urban mobility, reduced traffic congestion, and healthier
lifestyles.

We are interested in this topic because commuting behavior reflects the
complex interplay of socio-economic realities, urban design, and
environmental priorities, shaping both our daily lives and the future of
our planet. Understanding commuting behavior can help policymakers
design targeted interventions that promote sustainable and equitable
transport systems. For society, this research supports efforts to reduce
emissions, improve public health, and ensure accessible mobility for
all, while contributing valuable insights to the academic study of
transport behavior, particularly within the French context.

# Research question

Our research question explores how commuting behavior varies across
different socio-economic, geographic, and infrastructural factors in
France. This project aims to identify patterns and differences in
commuting choices based on key factors, such as urbanization, income,
population density, commuting distance, and accessibility to
transportation infrastructure, using large publicly available datasets
with varying levels of aggregation. By cleaning, harmonizing, and
analyzing these datasets, we will summarize variables and visualize
relationships to better understand how these factors correlate with
commuting behavior. While this study does not attempt to infer causal
relationships, it seeks to produce clear visualizations and highlight
meaningful patterns that can inform policymakers and contribute to
ongoing research on transportation behavior in France.

Here are some factors that might influence commuting behavior and the
associated variables that we have thought of:

-   **Socio-demographic factors**: age, gender, educational attainment
    and socio-professional category
-   **Income levels**: median living standard
-   **Infrastructures and urban/rural areas**: proximity to transport
    infrastructure and population density
-   **Commuting distances**: commuting destinations

Our project involves several steps: firstly, we collect, store, and
organize datasets in a GitHub repository. Since datasets are aggregated
at varying levels (e.g., “commune” vs. “intercommunal” and with
"lon/lat"), we then have to harmonize these differences using data
management methods seen in class. In parallel, we look for missing or
inconsistent data that may complicate our data analysis. After this, we
start initial data exploration, summarize variables using R and
ultimately generate visualizations to identify patterns.

# Datasets

Our analysis leverages multiple datasets from various reliable sources,
providing a comprehensive view of commuting behavior in France. From
INSEE's "Statistiques Locales" platform, we use demographic data on
population and density, income levels, and commuting patterns,
aggregated at intercommunal or municipal levels. These datasets offer
insights into variables such as population density, median income, car
ownership, and the use of different transport modes. Additionally, an
experimental dataset from transport.data.gouv.fr provides geolocated
information on public transport stops, enriching our understanding of
infrastructure accessibility. Further, the INSEE NAV2B dataset from the
2017 census highlights transport choices by employed individuals aged 15
and above, disaggregated by gender and geography. Lastly, the 2024
professional mobility dataset offers granular data on commuting
patterns, household vehicle ownership, and education levels, linking
residence and workplace at the municipal level. These diverse datasets,
despite differing levels of aggregation, collectively form the
foundation for our exploration of commuting behavior in France.

```{r}
library(vroom)
library(tidyverse)
library(here)
library(dplyr)
library(ggplot2)
library(tidyr)
here::i_am("Final_Project_Commuter_behavior.Rproj")

source <- "https://www.data.gouv.fr/fr/datasets/r/69cf54c6-6591-4920-b1d6-2a5292964606"
local_name <- "stops_france.csv.gz"
local_dir <- "data"
## is the directory available?
if (!dir.exists(here(local_dir))) {
  ## create the directory if needed
  dir.create(here(local_dir))
}
## is the file available?
if (!file.exists(here(local_dir, local_name))) {
  ## we need to download it
  stops_france <- vroom(source)
  ## let's save the file to avoid useless downloading
  vroom_write(stops_france, here(local_dir, local_name), delim = ",")
} else {
  ## we need only to load it
  stops_france <- vroom(here(local_dir, local_name))
}
```

```{r}
#Loading mob_pro
library(zip)
# Define variables
source <- "https://www.insee.fr/fr/statistiques/fichier/8205896/RP2021_mobpro.zip"
local_name <- "RP2021_mobpro.zip"
local_dir <- "data"
csv1_name <- "FD_MOBPRO_2021.csv"
csv2_name <- "varmod_mobpro_2021.csv"
# Check if the directory exists, create if not
if (!dir.exists(here(local_dir))) {
  dir.create(here(local_dir))
}
# Check if the ZIP file exists locally
if (!file.exists(here(local_dir, local_name))) {
  # Download the ZIP file
  download.file(source, destfile = here(local_dir, local_name), mode = "wb")
}
# Unzip the file
unzip(here(local_dir, local_name), exdir = here(local_dir))
# Load the first CSV file
mob_pro <- vroom(here(local_dir, csv1_name))
# Load the second CSV file
var_mob_pro <- vroom(here(local_dir, csv2_name))
```

```{r}
#Loading iltduu_sexe_trans

# Define variables
source <- "https://www.insee.fr/fr/statistiques/fichier/4515510/BTT_TD_NAV2B_2017.zip"
local_name <- "BTT_TD_NAV2B_2017.zip"
local_dir <- "data"
csv_name <- "BTT_TD_NAV2B_2017.CSV"
# Check if the directory exists, create it if not
if (!dir.exists(here(local_dir))) {
  dir.create(here(local_dir))
}
# Check if the ZIP file exists locally
if (!file.exists(here(local_dir, local_name))) {
  # Download the ZIP file
  download.file(source, destfile = here(local_dir, local_name), mode = "wb")
}
# Unzip the file
unzip(here(local_dir, local_name), exdir = here(local_dir))
# Load the specified CSV file
iltduu_sexe_trans <- vroom(here(local_dir, csv_name))
```

```{r}
#Loading the local files
demo <- vroom(here("data", "INSEE_Statistiques_locales_demographie.csv"))
revenus <- vroom(here("data","INSEE_Statistiques_locales_menages_fiscaux_niveau_de_vie.csv"))
local_mob <- vroom(here("data","INSEE_Statistiques_locales_territoires_ville_quartiers.csv"))
```

**Summary Table**

```{r echo = FALSE, results='asis'}
# Load required libraries
library(tibble)
library(knitr)

# Define datasets
datasets <- list(
  iltduu_sexe_trans = iltduu_sexe_trans,
  stops_france = stops_france,
  mob_pro = mob_pro,
  var_mob_pro = var_mob_pro,
  revenus = revenus,
  demo = demo,
  local_mob = local_mob
)

  
  # Create a summary table for all datasets
summary_table <- do.call(rbind, lapply(names(datasets), function(name) {
  data <- datasets[[name]]
  tibble(
    Dataset = name,
    `Number of Rows` = nrow(data),
    `Number of Columns` = ncol(data)
  )
}))

 # Print a heading and the table in LaTeX format
print(knitr::kable(summary_table, format = "pipe"))
```

# Data Analysis

With these datasets, we now aims to uncover patterns in commuting
behavior across France by exploring socio-demographic, economic,
geographic, and infrastructural variables. By systematically examining
these factors, we seek to identify trends and relationships that shape
commuting choices.

## Socio-demographic factors

Socio-demographic factors play a critical role in shaping commuting
behavior, and our hypotheses aim to explore how these factors influence
transport choices. These hypotheses, which align with common assumptions
in the field, suggest that younger individuals are more likely to use
active transport modes, such as cycling or walking, or rely on public
transport, while older populations may prefer private vehicles.
Similarly, men are hypothesized to have a higher likelihood of using
private vehicles or motorized two-wheelers, whereas women are expected
to rely more on public transport. We also posit that individuals with
higher educational attainment are more likely to choose sustainable
transport options, potentially due to greater environmental awareness or
because they tend to live in bigger cities, in which these modes of
transport are more common. Finally, socio-professional categories may
determine affordability and accessibility, with higher-income
professionals favoring private vehicles and individuals in intermediate
or lower-income categories depending more on public transport. This
section aims to verify these hypotheses by analyzing the relationships
between these socio-demographic characteristics and transportation
behavior.

### Age

```{r}
mob_age_aggregated <- mob_pro_cleaned %>%
  mutate(
    age_start = as.numeric(sub("-.*", "", age_five_year_slice)),
    age_10_year_slice = case_when(
      age_start >= 15 & age_start <= 24 ~ "15-24",
      age_start >= 25 & age_start <= 34 ~ "25-34",
      age_start >= 35 & age_start <= 44 ~ "35-44",
      age_start >= 45 & age_start <= 54 ~ "45-54",
      age_start >= 55 & age_start <= 64 ~ "55-64",
      age_start >= 65 & age_start <= 74 ~ "65-74",
      age_start >= 75 & age_start <= 84 ~ "75-84",
      age_start >= 85 & age_start <= 95 ~ "85-95",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(age_10_year_slice), age_start < 65) %>%  # Exclude age groups 65 and above
  group_by(age_10_year_slice, main_transportation_work) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(age_10_year_slice) %>%
  mutate(share = count / sum(count)) %>%  # Calculate share for each transportation mode in the age group
  ungroup()

```

```{r}
ggplot(mob_age_aggregated, aes(x = age_10_year_slice, y = share, fill = age_10_year_slice)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ main_transportation_work, scales = "free_y") +
  scale_y_continuous(labels = scales::percent_format()) +  # Show percentages on y-axis
  labs(title = "Share of Transportation Modes by Age Group (Below 65)", 
       x = "Age Group", 
       y = "Share of Transportation (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")  # Hide legend to reduce clutter
```

The graph shows the share that each transportation mode represents for
each age group. (For example, 4% of the 25 to 34 year-olds use their
bike to get to work.) It illustrates how the most common modes of
commuting to work vary across age groups and overall aligns with the
hypothesis that younger individuals prefer active and public transport
modes, while older individuals gravitate toward private vehicles. Public
transport usage is highest among the 15–24 age group and declines
steadily with age, reflecting its popularity among younger individuals.
Active transport modes show mixed trends: cycling decreases slightly
with age, while walking is prominent in both the youngest (15–24) and
oldest (55–64) groups. Private vehicle usage increases with age,
stabilizing among middle-aged groups, consistent with older individuals'
preference for comfort and convenience. Finally, the "no transport"
category rises sharply among those aged 55–64, indicating reduced
commutes in this group, maybe in cases of early retirement. Overall,
regardless of age, commuting by car is still the norm, followed by the
second most common transportation mode: public transport.

### Sex

```{r}
mob_gender <- mob_pro_cleaned %>%
  group_by(sexe, main_transportation_work) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(main_transportation_work) %>%
  mutate(share = count / sum(count)) %>%  # Calculate share for each gender within each transportation mode
  ungroup()

```

```{r}
ggplot(mob_gender, aes(x = sexe, y = share, fill = sexe)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ main_transportation_work, scales = "free_y") +
  scale_y_continuous(labels = scales::percent_format()) +  # Show percentages on y-axis
  scale_fill_manual(values = c("Male" = "#1f77b4", "Female" = "#ff7f0e")) +  # Set distinct colors for male and female
  labs(title = "Popularity of each Transportation Mode by Sex", 
       x = "Sex", 
       y = "Share of Transportation (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5),
        legend.position = "none")
```

This graph focuses on each transportation mode, indicating the sex that
primarily utilizes each one. (For instance 60% of the people, who
commute by bike, are males.) It highlights sex-based differences in
transportation preferences. Women rely more on public transport and
walking, supporting the hypothesis that they favor these modes for
commuting. Men exhibit a stronger preference for motorized two-wheelers
and bicycles, as well as a slightly higher reliance on cars/trucks/vans.
The "no transport" category is notably higher for women, possibly
indicating a greater proportion of non-commuters or shorter trips not
requiring traditional transport. These patterns align with the
hypothesis that women prefer walking and public transport, while men
lean toward private and motorized options, influenced by accessibility,
commuting needs, and societal roles.

### Educational attainment

```{r}
mob_education_aggregated <- mob_pro_cleaned %>%
  mutate(
    education_group = case_when(
      highest_qualification %in% c("No schooling", "Primary or lower", "Secondary or higher", "CEP", "BEPC") ~ "Group 1",
      highest_qualification %in% c("CAP/BEP", "General Bac", "Professional Bac") ~ "Group 2",
      highest_qualification %in% c("BTS/DUT", "Bachelor") ~ "Group 3",
      highest_qualification %in% c("Master", "Doctorate") ~ "Group 4",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(education_group)) %>%  # Exclude rows with missing education group
  group_by(education_group, main_transportation_work) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(education_group) %>%
  mutate(share = count / sum(count)) %>%  # Calculate share for each transportation mode in the education group
  ungroup()

```

```{r}
ggplot(mob_education_aggregated, aes(x = education_group, y = share, fill = education_group)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ main_transportation_work, scales = "free_y") +
  scale_y_continuous(labels = scales::percent_format()) +  # Show percentages on y-axis
  labs(title = "Share of Transportation Modes by Education Group", 
       y = "Share of Transportation (%)",
       x = "Education Group") +  # Explicit x-axis label
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5),  # Ensure x-axis text is centered
    legend.position = "bottom",  # Place legend at the bottom
    strip.text = element_text(size = 10),  # Adjust facet label size
    plot.title = element_text(hjust = 0.5),  # Center-align the title
    panel.spacing = unit(1, "lines")  # Add spacing between facets
  ) +
  guides(fill = guide_legend(title = "Education Group"))  # Add a clear legend title

```

We grouped the different levels of education in the following way:

*Group 1: No schooling, Primary or lower, Secondary or higher, CEP,
BEPC*

*Group 2: CAP/BEP, General Bac, Professional Bac*

*Group 3: BTS/DUT, Bachelor*

*Group 4: Master, Doctorate*

This graph illustrates the share of transportation modes used by
different education groups, categorized from Group 1 (lowest education
level) to Group 4 (highest education level). (For example, 70% of
people, whose highest qualification level is BTS/DUT or Bachelor,
commute by car, truck or van.) It highlights a correlation between
educational attainment and transportation preferences. Group 4,
representing those with the highest education, shows the highest share
of public transport and bicycle usage, reflecting potential
environmental awareness and most likely urban accessibility. In
contrast, reliance on cars/trucks/vans is highest in Groups 1 and 2.
Walking ("on foot") is relatively consistent across all groups, with a
slight increase in Group 4. These findings align with the hypothesis,
suggesting education influences awareness, affordability, and access to
sustainable transport.

### Socio-professional category

```{r}
mob_pro_cleaned_socio <- mob_pro_cleaned %>%
  filter(!is.na(socio_professional_cat), !is.na(main_transportation_work))  # Remove missing values

```

```{r}
mob_socio_prof_aggregated <- mob_pro_cleaned_socio %>%
  group_by(socio_professional_cat, main_transportation_work) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(socio_professional_cat) %>%
  mutate(share = count / sum(count)) %>%  # Calculate the share of each transportation mode
  ungroup()

```

```{r}
ggplot(mob_socio_prof_aggregated, aes(x = socio_professional_cat, y = share, fill = socio_professional_cat)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ main_transportation_work, scales = "free_y") +
  scale_y_continuous(labels = scales::percent_format()) +  # Display percentages on y-axis
  labs(
    title = "Share of Transportation Modes by Socio-Professional Category",
    x = "Socio-Professional Category",
    y = "Share of Transportation (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"  # Remove legend for clarity
  )

```

The different socio-professional categories can be translated as such:

*Agriculteurs exploitants (Farmers), Artisans (Craftsmen), Cadres
(Executives), Employés (Employees), Ouvriers (Workers), Professions
Intermédiaires (Intermediate Professions)*

This graph illustrates the distribution of transportation modes by
socio-professional category, highlighting key differences that align
with the hypothesis that socio-economic factors influence commuting
behavior. (For instance, 2% of craftsmen ("Artisans") commute using a
motorized two−wheelers.) Farmers ("Agriculteurs exploitants")
predominantly fall under "No Transport," likely reflecting their rural
settings and lack of regular commutes. Public transport is most
frequently used by executives ("Cadres") and employees ("Employés"),
possibly due to urban job locations and access to infrastructure.
Workers ("Ouvriers") show a high reliance on cars/trucks/vans,
reflecting affordability and accessibility in suburban or rural areas.
Motorized two-wheelers are notably used by craftsmen and workers, likely
due to flexibility and affordability. Bicycles and walking are
moderately used across categories, with a higher share among employees.
Overall, this supports the hypothesis that socio-professional categories
shape transport choices based on income, job type, and location.

## Income levels

Income levels significantly influence transportation choices by shaping
affordability and access to various modes. We hypothesize that wealthier
households favor private vehicles or motorized transport for their
convenience and comfort, while lower-income individuals rely more on
public transport or walking due to cost constraints. This section
investigates how income disparities impact commuting behavior and
whether observed patterns align with these assumptions, offering
insights into the role of economic factors in mobility decisions.

```{r}
# Load required libraries
library(dplyr)
library(ggplot2)
library(tidyr)

# Create income categories based on the median living standard
new_dataset$income_category <- cut(new_dataset$median_living_standard_2021, 
                                   breaks = quantile(new_dataset$median_living_standard_2021, probs = seq(0, 1, 0.25), na.rm = TRUE), 
                                   labels = c("Low", "Lower-Middle", "Upper-Middle", "High"), 
                                   include.lowest = TRUE)

# Sum transportation mode values for each income category
transport_columns <- c("no_transport_value", "on_foot_value", "bicycle_value",
                       "motorized_two_wheelers_value", "car_truck_van_value", "public_transport_value")

transport_by_income <- new_dataset %>%
  group_by(income_category) %>%
  summarise(across(all_of(transport_columns), sum, na.rm = TRUE))

# Normalize the data within each income category
transport_by_income_percent <- transport_by_income %>%
  rowwise() %>%
  mutate(total = sum(c_across(all_of(transport_columns))),
         across(all_of(transport_columns), ~ . / total * 100)) %>%
  select(-total) %>%
  pivot_longer(cols = starts_with("no_transport_value"):starts_with("public_transport_value"),
               names_to = "mode",
               values_to = "percentage")

# Plot the normalized data
ggplot(transport_by_income_percent, aes(x = income_category, y = percentage, fill = mode)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Transportation Modes by Income Category",
       x = "Income Category",
       y = "Percentage of Transportation Mode Usage",
       fill = "Transportation Mode") +
  theme_minimal() +
  theme(legend.position = "right")
```

This graph illustrates the relationship between income categories (Low,
Mid-Low, Mid-High, High) and transportation choices. To be more precise,
due to the availability of the data, this graph shows the share that
each transportation mode represents in each intercommunauté
(intercommunity), depending on the median living standard of that
intercommunity. Our observations are therefore less precise than if we
had conducted the analysis at the household level. (For example, in low
income intercommunities, 32% of people commute by public transport.)
Public transport usage is highest among the low-income group and
decreases as income rises, highlighting its affordability for
lower-income individuals. Similarly, walking ("on foot") and commuting
by motorized two-wheelers is more prevalent in lower-income
intercommunities, reflecting reliance on cost-effective transport modes.
Car/truck/van usage, represented by the dominant segment, remains high
across all income groups, suggesting widespread dependence on private
vehicles regardless of income. Bicycle usage is slightly in higher in
low income intercommunities, possibly due to urban accessibility and
environmental awareness. The "no transport" category doesn't seem to
vary across the different income categories. Overall, the graph aligns
with the hypothesis that higher income groups prioritize convenience,
while lower-income individuals favor affordable transport modes.

## Infrastructures and urban/rural areas

Proximity to transport infrastructure and urban/rural areas
significantly influence the accessibility and practicality of
transportation modes. Our hypothesis suggests that urban residents,
living in higher-density areas, are more likely to use public transport,
due to better infrastructure availability. Conversely, rural residents,
facing lower population densities and limited access to public
transport, are expected to rely more on private vehicles for their
commutes. This section examines how geographic and infrastructural
factors shape commuting behavior, offering insights into the disparities
between urban and rural mobility patterns and the implications for
sustainable transport planning. We will use population density to
estimate whether an intercommunity is urban or rural. To estimate the
quality of infrastructure we will use the number of stops from our
experimental dataset, keeping in mind that this data is very
approximative.

### Infrastructures

```{r}
# Scatterplot to visualize the relationship
library(ggplot2)

ggplot(final_dataset_cleaned2.0, aes(x = nb_stops, y = public_transport_share)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Relationship Between Public Transport Share and Number of Stops",
    x = "Number of Stops",
    y = "Public Transport Share"
  ) +
  theme_minimal()
```

This scatterplot shows the relationship between the number of public
transport stops and the share of public transport usage. The positive
trend, represented by the red regression line, indicates that areas with
a higher number of stops generally have a greater share of public
transport usage. This aligns with the hypothesis that proximity to
transport infrastructure enhances accessibility and encourages public
transport adoption. While the relationship is evident, the wide
dispersion of points, particularly at lower stop counts, suggests that
other factors, such as urban density or socio-economic characteristics,
also influence public transport usage. Overall, this graph supports the
idea that improved transport infrastructure, such as increasing the
number of stops, can facilitate greater reliance on public transport,
especially in densely populated urban areas.

### Uban/rural areas

```{r}
final_dataset_cleaned2.0_3 <- final_dataset_cleaned2.0 %>%
  mutate(urban_rural = ifelse(population_density_2021 > 150, "Urban", "Rural"))

# Boxplot for Urban/Rural vs. Car Share for Work
ggplot(final_dataset_cleaned2.0_3, aes(x = urban_rural, y = car_truck_van_share)) +
  geom_boxplot() +
  labs(
    title = "Car Share for Work by Urban/Rural Classification",
    x = "Urban/Rural",
    y = "Car Share for Work"
  ) +
  theme_minimal()
```

This boxplot compares car usage for work between rural and urban areas,
highlighting differences based on population density. The median car
share is higher in rural areas, with limited variation around the
central value, indicating a strong dependence on private vehicles. This
aligns with the hypothesis that rural residents rely more on cars due to
limited public transport options and greater travel distances. In urban
areas, while car use remains significant, the median share is lower, and
the distribution shows a wider range, reflecting greater availability of
alternative transport modes such as public transit or active modes.
Overall, the graph supports the hypothesis that urban residents are less
reliant on private vehicles compared to rural residents, largely due to
better transport infrastructure and shorter commute distances.

## Commuting distances

Commuting distance plays a critical role in determining the feasibility
and practicality of different transport modes. Shorter distances are
often associated with active modes such as walking or cycling, as they
require less time and physical effort. In contrast, longer distances
typically necessitate the use of private vehicles or public transport,
offering greater efficiency and comfort for extended commutes. For very
long commutes, individuals may rely on private cars or trains, depending
on the availability and convenience of infrastructure. This section
examines how commuting distances influence transportation choices,
shedding light on the relationship between distance and mode
feasibility, and providing insights into the dynamics of long- and
short-distance commuting behavior.

```{r}
mob_cleaned_distance <- mob_pro_cleaned %>%
  filter(!is.na(workplace_indicator))  # Remove rows with missing values in workplace_indicator

```

```{r}
mob_distance_aggregated <- mob_cleaned_distance %>%
  group_by(workplace_indicator, main_transportation_work) %>%
  summarize(count = n(), .groups = "drop") %>%
  group_by(workplace_indicator) %>%
  mutate(share = count / sum(count)) %>%  # Calculate the share of each transportation mode
  ungroup()

```

```{r}
ggplot(mob_distance_aggregated, aes(x = workplace_indicator, y = share, fill = workplace_indicator)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ main_transportation_work, scales = "free_y") +
  scale_y_continuous(labels = scales::percent_format()) +  # Show percentages on y-axis
  labs(
    title = "Share of Transportation Modes by Commuting Distance",
    x = "Commuting Distance (Workplace Indicator)",
    y = "Share of Transportation (%)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"  # Remove legend for clarity
  )

```

Commuting distance increases from 1 to 7, depending on workplace
indicator.

1.  Within the current commune of residence

2.  In another commune within the same department of residence

3.  In another department within the same region of residence

4.  Outside the current region of residence: in mainland France

5.  Outside the current region of residence: in an overseas department
    (DOM)

6.  Outside the current region of residence: in an overseas community
    (COM)

7.  Abroad

This graph illustrates the share of transportation modes by commuting
distance, correctly ordered from closest "Within Residence Commune" to
furthest "Abroad", as per the provided reference. (For example, almost
2% of the people working abroad bike to work.) Walking and cycling are
most prevalent for the shortest distances, such as "Within Residence
Commune," supporting the hypothesis that active transport is feasible
for closer commutes. Car usage dominates intermediate distances like
"Other Commune in Same Dept." and "Abroad". Public transport peaks for
longer distances, particularly for "Other Dept. (Same Region)".
Motorized two-wheelers are favored for commutes in "Outside Region:
COM". This interpretation confirms that commuting distances shape
transportation preferences effectively.\
*Note: when coding we should have ordered these workplace indicators
using "levels" for improved readability.*

# Discussion

## Summary of data analysis

Our analysis revealed several findings regarding commuting behavior in
France. From the regression models, population density emerged as a
significant determinant of transportation choices, with higher densities
strongly correlating with increased public transport usage and reduced
car dependency. Income levels also played a role, as lower-income groups
showed a higher reliance on public transport and active modes like
walking, while higher-income groups favored private vehicles.
Additionally, gender differences were shown, with men using motorized
two-wheelers more frequently and women relying slightly more on public
transport. Education levels were linked to sustainable transportation
choices, with higher-educated individuals showing greater shares of
public transport and bicycle usage. The analysis of commuting distances
also showed the practicality of certain transport modes, with public
transport usage increasing for longer commutes and active modes like
walking or cycling being dominant for shorter distances. These findings
show the links among socio-demographic, economic, geographic factors, as
well as commuting behaviors. While our study focuses on correlations
rather than causations, it tries to identify significant trends that
could inform or guide future research. \## The influence of our
variables Different factors likely play varying roles in influencing
commuting behavior. For instance, population density and access to
public transport infrastructure are expected to be more impactful than
individual-level variables like education. To better understand the
relative importance of these factors, we could apply a chi-squared (χ²)
test to examine the independence and association of categorical
variables (e.g., socio-professional category and transport modes).
However, we do recognize that the associations observed in this study
cannot account for confounding variables, such as individual
preferences, workplace policies, or social norms. A causal analysis
would require further access to public data and a rigorous regression
analysis. \## Modelling commuting behavior Using regression models, we
explored how variables like population density, income levels, and
public transport availability shape commuting choices. These models
provide insights into how certain factors—like proximity to transport
stops or urbanization—relate to increased public transport use, while
others, such as higher income, correlate with reliance on private
vehicles. Modelling could be expanded to predict commuting behavior
based on these variables, and could maybe be applied to policymaking, by
simulating the impact of interventions like expanding public transport
networks or incentivizing active modes of commuting (e.g., walking or
cycling). \## The role of remote working The COVID-19 pandemic brought
large changes to commuting patterns, with remote working reducing
overall commute volumes and altering transportation mode shares. It
would be interesting to extend our research by comparing pre- and
post-COVID commuting data to measure the long-term impact of remote
working on commute behavior. This could reveal shifts in preferences and
highlight the potential for permanent changes in urban mobility. \##
Possible future directions While our analysis identifies trends and
relationships, our study has several limitations, such as the inability
to infer causation and the potential for unmeasured confounding
variables. Integrating additional data sources, such as workplace
policies or real-time transport usage, could improve the robustness of
our findings. By better modeling commuting behavior and accounting for
changes like remote work, we could better understand how to design
policies that effectively promote sustainable and equitable
transportation systems.

# Conclusion

In the Harvard Business Review article "Why It’s So Hard to Change
People’s Commuting Behavior" (source 3) by Ariella Kristal and Ashley
Whillans, the authors emphasize how entrenched commuting habits are and
the difficulty in shifting them, even when presented with sustainable or
convenient alternatives. This underscores the significance of
understanding the factors that drive commuting choices, as any
meaningful change requires addressing deeply ingrained behaviors,
environmental constraints, and individual preferences. Our analysis
highlighted patterns of commuting behavior, such as the link between
population density and transportation choices. Income and education
levels also emerged as influential in our findings. Moreover, gender and
commuting distances revealed further differences. While our study
identified meaningful correlations, it emphasized the complexity of
commuting behavior and the need for further research to account for
causal relationships and confounding variables. Future work could
explore the long-term impacts of remote working, policy interventions,
and urban planning on commuting patterns.

# Sources

1.  https://www.statistiques.developpement-durable.gouv.fr/edition-numerique/chiffres-cles-transports-2023/20-emissions-de-gaz-a-effet#:\~:text=Les%20%C3%A9missions%20des%20transports%20a%C3%A9riens%20sont%20ainsi%20deux%20fois%20moins,de%20GES%20de%20la%20France.

2.  https://statistiques-locales.insee.fr/#view=map1&c=indicator

3.  https://hbr.org/2019/12/why-its-so-hard-to-change-peoples-commuting-behavior
